Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-05-16 ms/spaceness.tex
peter edits

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance. This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}. Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}), much modern work still describes geographic structure as a set of populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}. For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.

The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
that is much larger than the sample \citep{wakeley2003gene}. The key assumption is that the individuals of each generation are \emph{exchangeable}, so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}. If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways. For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations. Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.% It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.

\textcolor{red}{Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS). Previous work has found that discrete population structure can create false signatures of population bottlenecks when attempting to infer demographic histories from microsatellite variation \citep{Chikhi2010}, statistics summarizing the site frequency spectrum (SFS) \citep{Ptak2002,Stadler2009,Onge2012}, or runs of homozygosity in a single individual \citep{Mazet2015}. The increasing availability of whole-genome data has led to the development of many methods that attempt to infer detailed trajectories of population sizes through time based on a variety of summaries of genetic data \citep{Liu2015,Schiffels2014,Sheehan2013,Terhorst2016}. 
Because all of these methods assume that the populations being modeled are approximately randomly mating, they are likely affected by spatial biases in the genealogy of sampled individuals \citep{Wakeley1999}, which may lead to incorrect inference of population changes over time \citep{Mazet2015}. However, previous investigations of these effects have focused on discrete rather than continuous space models, and the level of isolation by distance at which inference of population size trajectories become biased by structure is not well known. Here we test how two methods suitable for use with large samples of individuals -- Stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016} -- perform when applied to populations evolving in continuous space with varying sampling strategies and levels of dispersal. }

@@

Last, some of the most basic tools of population genetics are summary statistics like $F_{is}$ and Tajima's $D$, which are often interpreted as reflecting the influence of selection or demography on sampled populations \citep{tajima1989statistical}. Statistics like Tajima's $D$ are essentially summaries of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals. Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972,Wakeley1999}, which can affect the value of Tajima's $D$ \citep{Stadler2009}. Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}. 
We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space, both to evaluate the assumptions underlying existing methods and to identify where the most promising signals of geography lie.

To study this, we have implemented an individual-based model in continuous geography 
that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and Drosophila, and output the full genealogy and recombination history of all final-generation individuals. 
We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
typically analyzed assuming discrete population models. 
We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
Last, we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

% The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
is determined by the interaction between rates of migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation Wright model is that this balance is nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species (see Discussion), but for reference estimates in human populations range from 40 to over 5,000 depending on the population and method of estimation (Table \ref{table:NStable}).
%\cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

The first approach to modeling continuously-distributed populations
was to endow individuals in a Wright-Fisher model with locations in continuous space.
However, since the total size of the population is constrained, this introduces interactions between arbitrarily distant individuals,
which (aside from being implausible) \cite{Felsenstein1975} showed would eventually lead to unrealistic population clumping
if the range is sufficiently large.
Another method for modeling spatial populations is to assume the existence of a grid of discrete randomly-mating populations connected by migration, 
thus enforcing regular population density by edict.
Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, 
\cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
Although these grid models are likely good approximations of continuous geography in many situations,
they do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
assumptions whose impacts are unknown. 
An alternative method for dealing with continuous geography is a new class of coalescent models,
the Spatial Lambda Fleming-Viot models \citep{Barton2010,Kelleher2014}.

@@

We simulated populations using the non-Wright-Fisher module in the program SLiM v3.0 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is equal to the mean parent-offspring distance.

@@

To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, 
we also implemented simulations identical to those above except that mates are selected uniformly random from the population,
and offspring disperse to a uniform random location on the landscape. 
We refer to this model as the ``random mating'' model, in contrast to the first, ``spatial'' model.

@@

%%%%%%%%%%%%%
\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. 
Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, 
which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples. The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}. For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (>48 map units) and nearby (<2 map units) individuals. Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance among individuals carrying each derived allele in a set of simulations representing a range of neighborhood sizes.

@@

%%%%%%%%%%%
\subsection{Demographic Modeling}
To assess the impacts of continuous spatial structure on demographic inference we inferred population size histories for all simulations using two approaches: Stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016}. Stairwayplot fits its model to a genome-wide estimate of the SFS, while SMC++ also incorporates information about linkage disequilibrium across the genome in a sequentially markovian coalescent framework. For both methods we sampled 20 individuals from all spatial simulations using random, midpoint, and point sampling strategies.

Stairwayplot documentation suggests fitting models across multiple bootstrap replicates drawn from empirical genomic data and taking the median inferred $N_{e}$ per unit time as the best estimate. We calculated site frequency spectra in scikit-allel \citep{Miles2017}, generated 100 bootstrap replicates per simulation by resampling over sites, and fit models for all bootstrap samples using default settings in stairwayplot.

For SMC++ we first converted simulated genotype matrices to VCFs with msprime and then used SMC++'s standard pipeline for preparing input files assuming no polarization error in the SFS. We used the first individual in the VCF as the "designated individual" when fitting models, and allowed the program to estimate the recombination rate during optimiziation. We fit models using the 'estimate' command rather than the now recommended cross-validation approach because our simulations had only a single contig.

To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy, and compared these values to the mean $N_{e}$ of random-mating simulations estimated from $\Theta_{\pi}$. We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model and plotting these against neighborhood size.

@@

To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with principal components as covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe \ak{need citation}. Conceptually our approach is similar to that taken in \cite{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero and standard deviation 10
to a mean that may depend on spatial position.
We adjust the geographic pattern of mean phenotype to reflect spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

We performed principal components analysis (PCA) using scikit-allel \citep{Miles2017} on the matrix of derived allele counts by individual for each simulation. 
SNPs were first filtered to remove strongly linked sites by calculating LD between all pairs of SNPs in a 200-SNP moving window and dropping one of each pair of sites with an $R^2$ over 0.1. The LD-pruned allele count matrix was then centered and all sites scaled to unit variance when conducting the PCA, following recommendations in \cite{Patterson2006}.

We ran linear-model GWAS both with and without the first 10 principal components as covariates in PLINK and summarized results across simulations by counting the number of SNPs 
with $p$-value below $0.05$ after adjusting for an expected false positive rate of less than 5\% \citep{benjamini2001control}. 
We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space. 
Results from all analyses were summarized and plotted with the 'ggplot2' \citep{Wickham2016} and ``cowplot'' \citep{Wilke2019} packages in R \citep{Rcore2018}.

@@

Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B). \textcolor{red}{Consistent with findings in n-island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation. This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).} 
Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B). Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.

@@

Values of observed heterozygosity and its derivative $F_{IS}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme. $F_{IS}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies. This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. Indeed, for random sampling under the spatial model, $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000. On the other hand, the dependency of raw observed heterozygosity on neighborhood size is not monotone. Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}). This squares with a report from \cite{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.

@@

\subsubsection{IBS tracts and correlations with geographic distance}
We next turn our attention to the effect of geographic distance on haplotype block length sharing, summarized for sets of nearby and distant individuals in Figure \ref{fig:ibs}. There are two main patterns to note. First, nearby individuals share more long IBS tracts than distant individuals (as expected because they are on average more closely related). Second, the difference in the number of long IBS tracts between nearby and distant individuals decreases as neighborhood size increases. This reflects the faster spatial mixing of populations with higher dispersal, which breaks down the correlation between the IBS tract length distribution and geographic distance. This can also be seen in the bottom row of Figure \ref{fig:sumstats}B, where the correlation coefficients between the summaries of the IBS tract length distribution (the mean, skew, and count of tracts over $1\times10^6bp$) and geographic distance approaches 0 as neighborhood size increases.

%Last, in figure \ref{fig:ibs} we can see that most of the difference both between nearby and distant individuals and across different neighborhood sizes is found in the upper tail of the IBS tract length distribution. This is consistent with the methods for inferring dispersal distances described in \cite{Ringbauer2017} and \cite{Baharian2016}, which focus on long IBS blocks rather than the full distribution.

>>>>

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance. This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}. Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}), much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}. For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.

The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
that is much larger than the sample \citep{wakeley2003gene}. The key assumption is that the individuals of each generation are \emph{exchangeable}, so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}. If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways. For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations. Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.% It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.

\textcolor{red}{Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS). Previous work has found that discrete population structure can create false signatures of population bottlenecks when attempting to infer demographic histories from microsatellite variation \citep{Chikhi2010}, statistics summarizing the site frequency spectrum (SFS) \citep{Ptak2002,Stadler2009,Onge2012}, or runs of homozygosity in a single individual \citep{Mazet2015}. The increasing availability of whole-genome data has led to the development of many methods that attempt to infer detailed trajectories of population sizes through time based on a variety of summaries of genetic data \citep{Liu2015,Schiffels2014,Sheehan2013,Terhorst2016}. 
Because all of these methods assume that the populations being modeled are approximately randomly mating, they are likely affected by spatial biases in the genealogy of sampled individuals \citep{Wakeley1999}, which may lead to incorrect inference of population changes over time \citep{Mazet2015}. However, previous investigations of these effects have focused on discrete rather than continuous space models, and the level of isolation by distance at which inference of population size trajectories become biased by structure is not well known. Here we test how two methods suitable for use with large samples of individuals -- stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016} -- perform when applied to populations evolving in continuous space with varying sampling strategies and levels of dispersal. }

@@

Last, some of the most basic tools of population genetics are summary statistics like $F_{IS}$ and Tajima's $D$, which are often interpreted as reflecting the influence of selection or demography on sampled populations \citep{tajima1989statistical}. Statistics like Tajima's $D$ are essentially summaries of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals. Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972,Wakeley1999}, which can affect the value of Tajima's $D$ \citep{Stadler2009}. Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}. 
We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space, both to evaluate the assumptions underlying existing methods and to identify where the most promising signals of geography lie.

To study this, we have implemented an individual-based model in continuous geography 
that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and Drosophila, and output the full genealogy and recombination history of all final-generation individuals. 
We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
typically analyzed assuming discrete population models. 
We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
Last, we examine the impacts of continuous geography on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between
migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation \citep[by][]{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
-- even in human populations,
estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

The first approach to modeling continuously-distributed populations
was to endow individuals in a Wright-Fisher model with locations in continuous space.
However, since the total size of the population is constrained, this introduces interactions between arbitrarily distant individuals,
which (aside from being implausible) 
was shown by \citet{Felsenstein1975} to eventually lead to unrealistic population clumping
if the range is sufficiently large.
Another method for modeling spatial populations is to assume the existence of a grid of discrete randomly-mating populations connected by migration, 
thus enforcing regular population density by edict.
Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, 
\citet{Rousset1997} showed that the slope of the linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
Although these grid models may be good approximations of continuous geography in many situations,
they do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
assumptions whose impacts are unknown. 
An alternative method for dealing with continuous geography is a new class of coalescent models,
the Spatial Lambda Fleming-Viot models \citep{Barton2010,Kelleher2014}.

@@

We simulated populations using the non-Wright-Fisher module in the program SLiM v3.1 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ is equal to the mean parent-offspring distance.

@@

To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, 
we also implemented simulations identical to those above except that mates are selected uniformly at random from the population,
and offspring disperse to a uniform random location on the landscape. 
We refer to this model as the ``random mating'' model, in contrast to the first, ``spatial'' model.

@@

%%%%%%%%%%%%%
\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. 
Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance \citep{Rousset1997}, 
which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes. The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}. For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (${}>48$ map units) and nearby (${}<2$ map units) individuals. Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance among individuals carrying each derived allele in a set of simulations representing a range of neighborhood sizes.

@@

%%%%%%%%%%%
\subsection{Demographic Modeling}

To assess the impacts of continuous spatial structure on demographic inference we inferred population size histories for all simulations using two approaches: stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016}. Stairwayplot fits its model to a genome-wide estimate of the SFS, while SMC++ also incorporates linkage information. % in a sequentially Markovian coalescent framework. 
For both methods we sampled 20 individuals from all spatial simulations using random, midpoint, and point sampling strategies.

As recommended by its documentation, we used stairwayplot to fit models with multiple bootstrap replicates drawn from empirical genomic data, and took the median inferred $N_{e}$ per unit time as the best estimate. We calculated site frequency spectra with scikit-allel \citep{Miles2017}, generated 100 bootstrap replicates per simulation by resampling over sites, and fit models for all bootstrap samples using default settings.

For SMC++, we first output genotypes as VCF with msprime and then used SMC++'s standard pipeline for preparing input files assuming no polarization error in the SFS. We used the first individual in the VCF as the ``designated individual'' when fitting models, and allowed the program to estimate the recombination rate during optimization. We fit models using the `estimate' command rather than the now recommended cross-validation approach because our simulations had only a single contig.

To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy.  We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model and plotting these against neighborhood size.

@@

To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with principal components as covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe \ak{need citation}. Conceptually our approach is similar to that taken by \citet{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero and standard deviation 10
to a mean that may depend on spatial position.
We adjust the geographic pattern of mean phenotype to create spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

We performed principal components analysis (PCA) using scikit-allel \citep{Miles2017} on the matrix of derived allele counts by individual for each simulation. 
SNPs were first filtered to remove strongly linked sites by calculating LD between all pairs of SNPs in a 200-SNP moving window and dropping one of each pair of sites with an $R^2$ over 0.1. The LD-pruned allele count matrix was then centered and all sites scaled to unit variance when conducting the PCA, following recommendations in \citet{Patterson2006}.

We ran linear-model GWAS both with and without the first 10 principal components as covariates in PLINK and summarized results across simulations by counting the number of SNPs 
with $p$-value below $0.05$ after adjusting for an expected false positive rate of less than 5\% \citep{benjamini2001control}. 
We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space. 
Results from all analyses were summarized and plotted with the ``ggplot2'' \citep{Wickham2016} and ``cowplot'' \citep{Wilke2019} packages in R \citep{Rcore2018}.

@@

Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B). \textcolor{red}{Consistent with findings in island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation. This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).} 
Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B). Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.

@@

Values of observed heterozygosity and its derivative $F_{IS}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme. $F_{IS}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies. This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. Indeed, for random sampling under the spatial model, $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000. On the other hand, the dependency of raw observed heterozygosity on neighborhood size is not monotone. Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}). This squares with a report from \citet{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.

@@

%%%%%%%%%%%%%%%
\subsubsection{IBS tracts and correlations with geographic distance}

We next turn our attention to the effect of geographic distance on haplotype block length sharing, summarized for sets of nearby and distant individuals in Figure \ref{fig:ibs}. There are two main patterns to note. First, nearby individuals share more long IBS tracts than distant individuals (as expected because they are on average more closely related). Second, the difference in the number of long IBS tracts between nearby and distant individuals decreases as neighborhood size increases. This reflects the faster spatial mixing of populations with higher dispersal, which breaks down the correlation between the IBS tract length distribution and geographic distance. This can also be seen in the bottom row of Figure \ref{fig:sumstats}B, where the correlation coefficients between the summaries of the IBS tract length distribution (the mean, skew, and count of tracts over $10^6$bp) and geographic distance approaches 0 as neighborhood size increases.

%Last, in figure \ref{fig:ibs} we can see that most of the difference both between nearby and distant individuals and across different neighborhood sizes is found in the upper tail of the IBS tract length distribution. This is consistent with the methods for inferring dispersal distances described in \citet{Ringbauer2017} and \citet{Baharian2016}, which focus on long IBS blocks rather than the full distribution.

The patterns observed for correlations of IBS tract lengths with geographic distance are similar to those observed in the more familiar regression of allele frequency measures such as $D_{xy}$ (i.e., ``genetic distance'') or $F_{ST}$ against geographic distance \citep{Rousset1997}. $D_{xy}$ is positively correlated with the geographic distance between the individuals, and the strength of this correlation declines as dispersal increases (Figure \ref{fig:sumstats}B), as expected \citep{Wright1943,Rousset1997}. This relationship is very similar across random and point sampling strategies, but is weaker for midpoint sampling, perhaps due to a dearth of long-distance comparisons. In much of empirical population genetics a regression of genetic differentiation against spatial distance is a de-facto metric of the significance of isolation by distance. The similar behavior of moments of the pairwise distribution of IBS tract lengths shows why haplotype block sharing has recently emerged as a promising source of information on spatial demography through methods described in \citet{Ringbauer2017} and \citet{Baharian2016}.
